Optimizing Abstract$^{\text 2}$ Machines

### 2012-11-07 paper/paper.tex
Filled in the details of uniform k-CFA

\begin{figure*}
\begin{align*}
\mathit{eval}(\mexp) &= \{ \mstate\ |\ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\\[2mm]
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mval,\msto}
\text{ where }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mlit,\msto}
\\
\ev{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep
\co{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto} & \machstep
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
\ap[^\mcntr_\mlab]{\mval,\mvalx{u},\mkont,\msto}
\text{ where }\mkont \in \msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep
\ev[^\mcntr]{\mexpi1,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\msto,\mvar,\mval)
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto}
\text{ where }\mkont\in\msto(\maddr)
\text{ and } \mval'\in\interpdelta(\mop,\mval)
\end{align*}
\caption{Abstract$^2$ machine for ISWIM}
\label{fig:aam}
\end{figure*}

\paragraph{Concrete interpretation} can be characterized by setting the implicit
parameters of the relation given in Figure~\ref{fig:aam} as follows:
\begin{align*}
\mathit{push}(\mlab,\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
\mbox{ where }\maddr \notin\msto
\\
\mathit{bind}(\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
\mbox{ where }\maddr \notin\msto
\end{align*}
The resulting relation is non-deterministic in its choice of
addresses, however it must always choose a fresh address when
allocating a continuation or variable binding.  If we consider machine
states equivalent up to consistent renaming, this relation defines
a deterministic machine.  (The relation is really a function.)

@@

\begin{align*}
\mathit{push} &= \dots\\
\mathit{bind} &= \dots
\end{align*}

\paragraph 0CFA

\begin{align*}
\mathit{push}(\mlab,\msto,\mkont) &= \mlab,\msto\sqcup[\mlab\mapsto\{\mkont\}]
\\
\mathit{bind}(\msto,\mvar,\mval) &= \menv[\mvar\mapsto\mvar],\msto\sqcup[\mvar\mapsto\{\mval\}]
\end{align*}

@@

The first four techniques reside in the purely functional realm. The
following three techniques use imperative constructs to avoid allocation
overhead.

@@

Lazy:
   cpu time: 32481 real time: 32881 gc time: 547

@@

@@

All the above techniques can be combined into a single implementation
of the small step reduction relation and small variations of a general
fixpoint combinator. We implementated all of these ideas in an
analysis of R4RS scheme in just 3000 lines of Racket. Most lines are
spent listing all the primitives and library functions (and their
abstract and concrete implementations) of R4RS. Here we describe the
unifying vocabulary for writing such a framework, and some design
decisions related to scaling to a language with compound data.

@@

The linguistic elements described above were all reusable to implement
the various languages' abstract interpreters. One such language was
R4RS Scheme, which has compound data (e.g. vectors and conses).

@@

There are a few steps to consider:
\begin{itemize}
 \item{Define special value lattice elements for compound data domains that can be quoted
       (e.g. {\tt QPair} for $\left\lbrace ({\tt cons} a_i d_i)\right\rbrace_i$, {\tt QVector} for immutable vectors, etc.)}
 \item{Define a ``larger'' value lattice element for all quotable data, {\tt QData}}
 \item{Interpret {\tt (quote (a ...))} as {\tt (qlist a ...)}, a new primitive function defined in figure \ref{fig:qlist},
       and similar definitions for immutable vectors.}
 \item{Extend the $\interpdelta$ axioms to include conservative meaning for these new values (e.g. {\tt (car QData)} = {\tt QData}, {\tt (add1 QData)} = {\tt Number} and log ``possible type error'') and allow them to allocate addresses and change the store}
\end{itemize}

\begin{figure}
\begin{align*}
\interpdelta(\msto, qlist) &= (\mbox{{\tt '()}}, \msto) \\
\interpdelta(\msto, qlist, v ..._+) &= (({\tt cons}\ a\ d), \msto') \\
\qquad \mbox{where } \msto' &= \msto\sqcup[a \mapsto \bigsqcup(v ...)] \\
                            &\qquad \sqcup[d \mapsto \lbrace \mbox{{\tt '()}}, ({\tt cons}\ a\ d)\rbrace] \\
\bigsqcup(v) &= v \\
\bigsqcup(v, vs ...) &= merge(v, \bigsqcup(vs ...)) \\
merge(v, v) &= v \\
merge(n, m) &= {\tt Number} \\
merge(n, v) &= {\tt QData} \\
merge(({\tt cons}\ a\ d), ({\tt cons}\ a'\ d')) &= {\tt QPair} \\
merge({\tt QPair}, ({\tt cons}\ a\ d)) &= {\tt QPair} \\
merge({\tt QPair}, v) &= {\tt QData} \\
\vdots
\end{align*}
\caption{Quoted list primitive}
\label{fig:qlist}
\end{figure}

>>>>

\begin{figure*}
\begin{align*}
\mathit{eval}(\mexp) &= \{ \mstate\ |\ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\\[2mm]
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mval,\msto}
\text{ where }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mlit,\msto}
\\
\ev{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep
\co{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto} & \machstep
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
\ap[^\mcntr_\mlab]{\mval,\mvalx{u},\mkont,\msto}
\text{ where }\mkont \in \msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep
\ev[^\mcntr]{\mexpi1,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\mvar,\mval)
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto}
\text{ where }\mkont\in\msto(\maddr)
\text{ and } \mval'\in\interpdelta(\mop,\mval)
\end{align*}
\caption{Abstract$^2$ machine for ISWIM}
\label{fig:aam}
\end{figure*}

\paragraph{Concrete interpretation} can be characterized by setting the implicit
parameters of the relation given in Figure~\ref{fig:aam} as follows:
\begin{align*}
\mathit{push}_\mlab^\mcntr(\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
\mbox{ where }\maddr \notin\msto
\\
\mathit{bind}(\menv,\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
\mbox{ where }\maddr \notin\msto
\end{align*}
The resulting relation is non-deterministic in its choice of
addresses, however it must always choose a fresh address when
allocating a continuation or variable binding.  If we consider machine
states equivalent up to consistent renaming, this relation defines
a deterministic machine.  (The relation is really a function.)

@@

\begin{align*}
\mathit{push}_\mlab^\mcntr(\msto,\mkont) &=
  \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
\mathit{bind}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto \maddr],
                                           \msto\sqcup[\maddr \mapsto
                                             \{ \mval\}],
                                           \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= x\mcntr' \\
%%              \lfloor \mtcntr \rfloor_k &= \mtcntr \\
              \lfloor \mcntr \rfloor_0 &= \mtcntr \\
              \lfloor \mlab\mcntr \rfloor_{k+1} &= \mlab\lfloor \mcntr\rfloor_k
\end{align*}
%% Redundant
%% \paragraph 0CFA

%% \begin{align*}
%% \mathit{push}(\mlab,\msto,\mkont) &= \mlab,\msto\sqcup[\mlab\mapsto\{\mkont\}]
%% \\
%% \mathit{bind}(\menv,\msto,\mvar,\mval) &= \menv[\mvar\mapsto\mvar],\msto\sqcup[\mvar\mapsto\{\mval\}]
%% \end{align*}

@@

We discuss a variety of techniques for increasing performance by
decreasing the interpretive overhead of abstract interpretation, all
in the purely functional realm. These techniques are the following:
\begin{itemize}
 \item{shrink state space by quotients (lazy non-determinism)}
 \item{shrink state space by removing {\tt ev} states (abstract
   compilation)}
 \item{decrease intermediate data structures by specializing our
fixpoint operator to work with widened machine configurations
(fixpoint specialization)}
 \item{decrease intermediate data structures by operating on store
   changes rather than whole stores (store deltas)}
\end{itemize}
Each technique requires minimal changes to the reduction
relation/fixpoint, and each improves performance by integer factors.

@@

The first problem that we realized with the AAM approach was that much
of the non-determinism is spurious. For example, in a function
application,
\begin{alltt}
(f x y z w)
\end{alltt}
where all variables have been joined with multiple values, each
variable reference transitions to several new states that comprise the
cartesian product of all the values of {\tt x}, {\tt y}, {\tt z} and
{\tt w}, only to be joined back together in {\tt f}'s bindings. For
this reason, we {\it delay} the non-determinism until it is needed in
{\it strict contexts} (such as the guard for an {\tt if} a called
procedure, or a numerical primitive application).

For this, we introduce a new value for a delayed lookup that acts to
quotient the state space by the values at the delayed address.

\begin{align*}
force(\clos{x,e,\menv},\msto) &= \{\clos{x,e,\menv}\}\\
force(\saddr\maddralt,\msto) &= \msto(\maddralt)\\
\ev{\svar{x},\menv,\mkont,\msto} &\machstep\; 
\co{\mkont,\saddr{\menv(x)},\msto} \\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},v,\msto}
&\machstep\;
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{v,\maddr}} \\
\mbox{where } &v' \in force(v,\msto)
bind^\mcntr_\mlab(\msto,x,\menv,v) =
\end{align*}
Lazy:
   cpu time: 32481 real time: 32881 gc time: 547

@@

\section{Sacrificing precision for performance}

Each step in the fixpointer for the above techniques requires joining
several (possibly large) stores together. Even with immutable hashes
that utilize sharing, this is memory-intensive. Worse, this
contributes an obscene fraction of the runtime. We reuse a technique
created in Shivers' dissertation~\cite{ianjohnson:Shivers:1991:CFA},
timestamp approximation, that removes the need for joining whole
stores by manipulating a global store imperatively. This loses
precision because the starting store for each state in a machine
configuration can change between stepping all states and introduce
more flows than originally possible. The example in figure
\ref{fig:timestamp} illustrates this loss in precision. {\bf TODO:}
EXPLANATION OF LOSS. In many applications, this is acceptable for the
performance improvement.

\begin{figure}
TODO
\caption{Precision lossy program in timestamp approximation}
\label{fig:timestamp}
\end{figure}

@@

All the above techniques can be combined into a single implementation
of the small step reduction relation and simple variations of a
general fixpoint combinator. We implementated all of these ideas in an
analysis of core Scheme (call-by-value lambda calculus extended with
variable-arity functions, call/cc, vectors, conses, recursive binding,
various primitives) in just 3000 lines of Racket. Most lines are spent
defining the primitives and expected library functions used in our
benchmarks. Here we describe the unifying vocabulary for writing such
a framework, and some design decisions related to scaling to a
language with compound data.

@@

The linguistic elements described above were all reusable to implement
the various languages' abstract interpreters. The language of our
benchmarks, Scheme, has compound data (e.g. vectors and conses), which
introduces choices for abstraction.

@@

There are a few steps to consider:
\begin{itemize}
 \item{Define special value lattice elements for compound data domains that can be quoted
       (e.g. {\tt QPair} for $\left\{ ({\tt cons} a_i d_i)\right\}_i$, {\tt QVector} for immutable vectors, etc.)}
 \item{Define a ``larger'' value lattice element for all quotable data, {\tt QData}}
 \item{Interpret {\tt (quote (a ...))} as {\tt (qlist a ...)}, a new primitive function defined in figure \ref{fig:qlist},
       and similar definitions for immutable vectors.}
 \item{Extend the $\interpdelta$ axioms to include conservative meaning for these new values (e.g. {\tt (car QData)} = {\tt QData}, {\tt (add1 QData)} = {\tt Number} and log ``possible type error'') and allow them to allocate addresses and change the store}
\end{itemize}

\begin{figure}
\begin{align*}
\interpdelta(\msto, qlist) &= (\mbox{{\tt '()}}, \msto) \\
\interpdelta(\msto, qlist, v ..._+) &= (({\tt cons}\ a\ d), \msto') \\
\qquad \mbox{where } \msto' &= \msto\sqcup[a \mapsto \bigsqcup(v ...)] \\
                            &\qquad \sqcup[d \mapsto \{ \mbox{{\tt '()}}, ({\tt cons}\ a\ d)\}] \\
\bigsqcup(v) &= v \\
\bigsqcup(v, vs ...) &= merge(v, \bigsqcup(vs ...)) \\
merge(v, v) &= v \\
merge(n, m) &= {\tt Number} \\
merge(n, v) &= {\tt QData} \\
merge(({\tt cons}\ a\ d), ({\tt cons}\ a'\ d')) &= {\tt QPair} \\
merge({\tt QPair}, ({\tt cons}\ a\ d)) &= {\tt QPair} \\
merge({\tt QPair}, v) &= {\tt QData} \\
\vdots
\end{align*}
\caption{Quoted list primitive}
\label{fig:qlist}
\end{figure}
